# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

The working data is already downloaded in the working directory, so in the next chunk of code will only be unzipped and read in R.

```{r}
library(plyr)
#Unzip the data file and read it, then change the date to Date-class type
unzip("activity.zip")
data<-read.csv("activity.csv")
data$date<-as.Date(data$date)

```
## What is mean total number of steps taken per day?

The chunk of code below make a histogram of the steps per day, showing the distribution that the steps have. The histogram show that 0 is a dominant value in the data set. This can be caused by all the values near to 0 or 0 that appear on the first intevals of time each day, that are produced because the people are sleeping (probably) at that intervals(0 to 5 am each day)

```{r,results='asis'}
#Create a new data frame without NA's
completedata<-data[complete.cases(data),]
#Make a histogram with the frequency of number of steps
hist(completedata$steps,main="Distribution of steps taken per day",xlab="Number of steps taken per day", ylab="Frequency",col="Red")

```

Next is showed a table with the mean of steps taken per day and the median. The median is in line with the histogram, showing that 0 is a dominant value.

```{r,results='asis'}
#Calculate the mean and the median of steps taken per day,assign names to the data frame 
meanmediantable<-ddply(completedata,.(date),summarise,mean(steps),median(steps))
#Assign names the columns of the data frame
names(meanmediantable)<-c("date","mean","median")
#Transform date variable to a character class variable to show correctly on table
meanmediantable$date<-as.character(meanmediantable$date)
#Prints the table in html format
library(xtable)
xt<-xtable(meanmediantable)
print(xt,type="html")
```


## What is the average daily activity pattern?

This section shows the pattern on time intervals, showing first a plot of the mean of steps by interval vs interval. This plot shows that in the first intervals the people don't move a lot, and the values are near to 0. The average number of steps have a huge boost beyond the 500 interval, having the highest peak at the interval 835. This effect on the plot can be explained because at 5:00 the people begins their daily activities, and at 8:35 the people are doing activities that requires more movement than in other intervals. After that point, the activity have a considerable decrease, having other peaks but not with lowest values than the 835 interval. This can be explained because the people in the data set are more stable in one place, like their work or school. Finally, the activity have a decreasing tendency after the 2000 interval. This can be explained because at that time, people are at home resting. 

```{r}
prueba<-ddply(data,.(interval),summarize,mean=mean(steps,na.rm=TRUE))
plot(prueba$interval,prueba$mean,type="l",xlab="Interval",ylab="Number of steps")
##Computes the interval with the maximum number of steps
max<-prueba[which.max(prueba$mean),]
```

The interval that contains, on average, more steps is `r max[1,1]` which have `r max[1,2]` steps

## Imputing missing values

The next chunk of code will count the number of rows where data was not available (NA). To avoid the NA's, a new data set is created replacing the NA values with the mean of the time interval where the NA value is contained. 

```{r}
#Counts the number of NA's in the data set.
nasteps<-sum(is.na(data$steps))
#Create a data set copying the original data set first.
filleddata<-data
#Create a boolean vector containing the rows where are located NA values.
na<-is.na(filleddata$steps)
#Matches the interval where the NA values are located with the mean computed on a chunk before.
matches<-match(filleddata$interval[na],prueba$interval)
#Replaces the NA value with the mean of the time interval.
filleddata$steps[na]<-prueba[matches,2]
```

The next chunk of code make a histogram of the steps taken per day with the filled data. As it can be seen, the general form of the histogram doesn't change  in comparison with the histogram made before, which doesn't consider the NA values. So, the NA values doesn't change at all the distribution of the number of steps taken per day, keeping the 0 as the dominant value, as the median of each time interval shows.

```{r}
#Make a histogram with filled data
hist(filleddata$steps,main="Distribution of steps taken per day",xlab="Number of steps taken per day", ylab="Frequency",col="Red")
```

The next chunk of code creates a table with the mean and the median of the number of steps. The median computed is in line with the histogram, showing that 0 is a dominant value as can be seen on the histogram.

```{r,results='asis'}
fillmeanmedian<-ddply(filleddata,.(date),summarise,mean(steps),median(steps))
names(fillmeanmedian)<-c("date","mean","median")
fillmeanmediantable<-xtable(fillmeanmedian)
print(fillmeanmediantable,type="html")
```

## Are there differences in activity patterns between weekdays and weekends?

In this section is analyzed the differences between weekdays and weekends activity. In the panel plot can be visualized both categories of days. In general, the form of both plots are very similar, having both plots peaks on the same intervals.The similarities are that both plots present highest values of steps beyond the 500 interval. Also both have a decreasing tendency beyond the 2000 interval. 
The main differences are that weekends days have highest peaks than the weekdays, but the highest value of steps is on a weekday somewhere between the 500 and 1000 interval. 
With the information described above, we can conclude that in general, there is more activity registration on weekends than in weekdays, probably because the people measuring their steps tends to stay out of their homes on weekends, causing to walk more.

```{r}
library("lattice")
#Create a new variable in the data set that is the weekday of the date
data$weekday<-weekdays(data$date)
#Create a table to look up the values of the days, and match them with weekends or weekdays.
days<-c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")
fact<-c(rep("weekday",5),rep("weekend",2))
weekdaystable<-cbind(days,fact)
#Matches the weekday of the data set with the look up table, to assign a value to the new variable of the data set named fact.
weekdaymatches<-match(data$weekday,weekdaystable[,1])
data$fact<-weekdaystable[weekdaymatches,2]
#Creates a data frame with the mean of the steps by interval and by type of day (weekend or weekday)
meantable<-ddply(data,.(interval,fact),summarize,mean=mean(steps,na.rm=TRUE))
#Make a panel plot, on the top the weekend inteval vs average steps plot, on the bottom the weekdays interval vs average steps plot.
xyplot(meantable$mean~meantable$interval|meantable$fact,layout=c(1,2),type="l",xlab="Interval",ylab="Number of Steps")
```



